---
title: "Difference tests and effect sizes with behavioural inhibition data"
geometry: margin=0.8in
output:
  pdf_document: default
  html_document: default
---

## Return to the mood experiment

We will use the same 58 participants from Paál et al. (2015) as in the earlier
practicals. Each person completed a stop-signal task after either a neutral or
negative mood induction. A longer stop-signal reaction time (SSRT) means slower
inhibition. The two conditions contain *different people*, so this is a
between-participants comparison.

```{r message=FALSE, warning=FALSE}
library(tidyverse)

d <- read_csv(
  "https://dfzljdn9uc3pi.cloudfront.net/2015/964/1/data_supplement.csv",
  show_col_types = FALSE
) %>%
  rename(Participant = 1) %>%
  mutate(Condition = factor(
    Mood_induction_condition,
    levels = c(2, 1),
    labels = c("Neutral", "Negative")
  ))

d %>% select(Participant, Age, Condition, SSRT) %>% slice_head(n = 6)
```

First, look at the observations and group summaries. These tell us about the
sample, but do not yet tell us how uncertain the difference is.

```{r fig.height=2.4}
ggplot(d, aes(x = Condition, y = SSRT)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.12, height = 0, alpha = 0.6) +
  labs(x = "Mood condition", y = "SSRT (ms)")
```

Now let's look at summary statistics again.

```{r}
group_summary <- d %>%
  group_by(Condition) %>%
  summarise(n = n(), mean_SSRT = mean(SSRT), sd_SSRT = sd(SSRT),
            .groups = "drop")
group_summary
```

### Practice 1

Calculate the differences between means, i.e. Negative - Neutral using the 
means estimated above. What can you tell about the difference? What are the
units that it is measured in?

```{r}
### YOUR CODE HERE ###
```

## A difference test is a linear model

The model is

$$SSRT_i = \beta_0 + \beta_1 Condition_i + \epsilon_i,$$

where `Condition` is 0 for Neutral and 1 for Negative. Thus $\beta_0$ is the
Neutral mean and $\beta_1$ is **Negative minus Neutral**. Under the null
hypothesis, the population difference is zero: $H_0: \beta_1 = 0$.

```{r}
m_mood <- lm(SSRT ~ Condition, data = d)
summary(m_mood)
```
We can also look at the confidence intervals.


```{r}
confint(m_mood)
```


### Practice 2

Compare the model's intercept and condition coefficient with the means and
difference from Practice 1. Read the coefficient's standard error and compute
its $t$ value by hand using

$$t = \frac{\text{estimated difference} - 0}{SE(\text{estimated difference})}.$$

Check your answer against the `t value` column above. Did you get the same value?

```{r}
### YOUR CODE HERE ###
```

## From a t value to a p value

For this two-sided test, the p value is the probability, **assuming a zero
population difference**, of a $t$ value at least as far from zero as the one
observed. It is not the probability that the null hypothesis is true. 

`pt(x, df)` gives the probability to the *left* of `x` on a $t$ distribution.
We can use the negative absolute value to obtain one tail, then double it for
a two-sided p value.

```{r}
t_observed <- summary(m_mood)$coefficients["ConditionNegative", "t value"]
df <- nrow(d) - 2
2 * pt(-abs(t_observed), df = df)
```

### Practice 3

What are the degrees of freedom and the two-sided p value? At a 0.05
threshold, would you reject $H_0$? Does this result establish that the two
population means are equal? Explain briefly.

For a direction chosen *before seeing the data*, calculate the one-sided p
value for $H_1$: Negative mood leads to longer mean SSRT. How does it
relate to the two-sided value here?

```{r}
### YOUR CODE HERE ###
```

## The same test without writing a model

For two independent groups, `t.test()` can test a difference in means. Here
`var.equal = TRUE` requests the same pooled-variance used by `lm()`.
`t.test()` subtracts the *second* factor level from the first in its printed
order (Neutral minus Negative here), so its $t$ value has the opposite sign
from `ConditionNegative`.

```{r}
t.test(SSRT ~ Condition, data = d, var.equal = TRUE)
```

### Practice 4

Compare the absolute $t$ value, degrees of freedom, p value, and 95%
confidence interval with the model results. Reverse the interval's signs to
express it as Negative minus Neutral. Does the interval include zero?

What happens if you say that variance is different? `var.equal = FALSE`.

```{r}
### YOUR CODE HERE ###
```

## Cohen's d: the size of the difference

A p value alone does not tell us how large a difference is. Cohen's $d$
expresses the mean difference in pooled standard-deviation units:

$$s_p = \sqrt{\frac{(n_0-1)s_0^2 + (n_1-1)s_1^2}{n_0+n_1-2}},
\qquad d = \frac{\bar{x}_{Negative}-\bar{x}_{Neutral}}{s_p}.$$

The weights matter here because the groups contain 28 and 30 people. $d$ has no
units; the unstandardised difference remains useful because it is in milliseconds.

### Practice 5

Use `group_summary` to calculate $s_p$ and Cohen's $d$. What does the sign
mean? Report both $d$ and the difference in milliseconds. Which measure is
easier to interpret?

```{r}
### YOUR CODE HERE ###
```


**Source.** The dataset is from Paál, Carpenter, and Nettle (2015),
[*Childhood socioeconomic deprivation, but not current mood, is associated
with behavioural disinhibition in adults*](https://doi.org/10.7717/peerj.964).
